Use and acquisition of causal relations for decision making under uncertainty using imperfect information games

### 2018-08-24 Propuesta/Formato/Proposal_format.tex
ultima version

% Title page
\input{gfrontpage}
%\doublespacing
\singlespacing
\tableofcontents
\newpage
\begin{abstract}
We consider decision problems under uncertainty where the options available to a decision maker and the resulting outcome are related through a causal mechanism which is unknown to the decision maker, although he is aware of the causal nature of his environment. We study how a decision maker can learn about this causal mechanism through sequential decision making as well as using current causal knowledge inside each round in order to make better choices had he not considered causal knowledge. We propose a decision making procedure in which an agent holds \textit{beliefs} about her environment which are used to make a choice and then are updated using the observed outcome. As proof of concept, we present an implementation of this causal decision making model and apply it to a simple problem. We show that the model achieves a performance similar to the classic Q-learning while it also acquires a causal model of the environment. 
\end{abstract}
\section{Introduction}
\indent A fundamental part of intelligent reasoning is being able to make decisions under uncertain conditions (\cite{danks2014unifying}, \cite{lake2017building}, \cite{pearlwhy}). Uncertain conditions mean that given a set of available choices, a decision maker can not predict the effect of the his choices. Because of this, the decision maker must take into consideration the stochastic relation between actions and consequences. From the von Neumann-Morgenstern decision making theory we know that when a rational\footnote{To be explained in full detail later.} decision maker knows this stochastic relation, then he must\textit{If rationality is assumed} choose the action that maximizes the \textit{expected utility} of a \textit{utility function} whose existence is guaranteed (\cite{von1944theory}). On the other hand, when a rational decision maker does not know the probabilities of the ocurrence of certain outcomes, or consequences, given his actions, then he must encode his personal, subjective, quantifications of any uncertainties in a probability distribution $p$ and choose the action that maximizes expected utility with respect to this probability (\cite{savage1954the}, \cite{bernardo2000bayesian}, \cite{gilboa2009decision}). If the current \textit{knowledge} of a decision maker is not enough to make good choices, then the decision maker can \textit{learn} from the environment by interacting with it in order to update his quantifications of uncertainty (\cite{bernardo2000bayesian}, \cite{peterson2017introduction}).

@@

Learning by interaction has been extensively studied by computer scientists using the Reinforcement Learning (RL) setting (\cite{sutton1998reinforcement}), but the most common used techniques  in this field are purely associative and do not consider any high-level structure of the environment beyond what is expressable in a Markov Decision Process (\cite{garnelo2016towards}). A particular case of a \textit{higher level structure}; i.e., beyond associative patterns, is the case of \textit{causal} structure. A causal structure encondes a series of \textit{cause-effect} relations between events and knowing such relations allows a decision maker to add extra knowledge into the uncertainty of his environment and also allows to plan ahead his actions since he can predict what a certain action will cause (\cite{spirtes2000causation}, \cite{pearl2018theoretical}).

Since human beings are known to learn causal models in sequential decision making processes (\cite{sloman2006causal}, \cite{nichols2007decision}, \cite{meder2010observing}, \cite{hagmayer2013repeated}, \cite{danks2014unifying}), and even though this learning is not perfect (\cite{rottman2014reasoning}), my hypothesis is that an autonomous agent can learn and use causal information while interacting with an uncertain environment which is governed by a fixed \textit{causal mechanism} which is unknown to the agent.

@@

Learning a causal model of a certain environment would allow transferability of knowledge into similar domains because the underlying cause-effect structure has been captured and could later be used.
\subsection{Outline}
In the Section \ref{Main_Concepts} the main concepts that serve as a foundation for this work will be described; in particular, we will describe in detail what causality means. Since this research proposal asumes that a causal model governs an environment, a precise definition of causality will be given along with the mathematical machinery to perform causal inference. Also, since the entity that will be facing causal environments is a \textit{rational} decision maker we will precisely define, axiomatically, what does rationality means and which are the main theories of rational decision making. Finally, since interactive learning is considered, we will describe a mathematical theory which is suitable to modelling interaction between rational decision makers.

@@

	Notice that since the graph is finite, there exist some nodes that does not have causes, which are called \textit{exogenous}. If an $A$ event is caused by some other event, then we say it is \textit{endogenous} and we denote the set of its causes as $Pa(A)$.

@@

		\subsubsection{Extensions}
		Given the Limitations addressed in the previous section, some possible extensions could be:
		\begin{itemize}
		\item Consider longer games.
		\item Consider more players than an agent and Nature.
		\item Consider objectives/intentions for Nature.
		\item Studying game-theoretical properties of the model such as equilibria in terms of the causal structure.
		\end{itemize}

Reinforcement Learning (RL) has been the standard setting for decision-making and learning by interaction. In this setting it is assumed that a Markov Decision Process (MDP) is able to capture the effects of the agent's actions in the environment. An MDP is composed by a set of states, a transition function and a reward function, where an agent's action moves the state of the environment according to the transition function and obtaining a reward in such process.

@@

\newpage
\section{Methodology}
In this section we describe the steps that are to be followed in order to answer the Hypothesis.
\subsection{A guiding principle for a Causal Decision Problem}
We propose the following principle as a normative guide for how causal information must be used in order to find a solution to a decision problem that is rational. We consider only the case where the decision maker's utility is given by the outcome of a $\{0,1 \}$-binary variable.
\begin{prop}
Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Let $Y$ a variable in $\mathcal{G}$ the variable whose realizations correspond to the consequences of the decision problem once the agent has chosen an action $a \in \mathcal{A}$ and assume that $Y$ only takes values in the set $\{ 0,1\}$ where $1$ is a more desired outcome for the agent. Then, the $\textit{solution}$ for this decision problem is given by the action $a^\ast \in mathcal{A}$ that satisfies:
\[ P(Y=1 | do(a^\ast) \geq P(Y=1 | do(a)) \textrm{ for all } a \in \mathcal{A}. \]
\end{prop}
\subsection{Modelling a Decision Problem under Uncertainty as a Game}
Consider a decision problem under uncertainty $(\mathcal{A},\mathcal{E},\mathcal{C},\succeq)$ where the available actions $\mathcal{A}$ to a decision maker are causally related to the outcomes $\mathcal{C}$ and this causal relations are captured by a Causal Graphical Model $\mathcal{G}$. The Causal model is assumed to remain fixed and unknown to the decision maker, who is assumed to be rational and aware of the causal nature of the decision problem he faces.

@@

If the maximum number of variables is assumed to be known by the decision maker, then the situation simplifies since a decision maker could generate distributions that represent causal relations among variables. For example, a Dirichlet Process (\cite{ferguson1973bayesian}, \cite{ghosal2017fundamentals}) could be used to generate Dirichlet distributions that are used generate Conditional Probability Tables in order to specify a causal model.

@@

\subsection{Balancing exploitationd and exploration}
When the model is to be discovered, such as in the general case mentioned before, an agent must choose also between keep making an intervention in a variable that is known to cause a desired output or to explore for new variables that could bring even better rewards, but at the cost of not having best performance while looking out for them; this is known as the exploration vs exploitation dilemma (\cite{sutton1998reinforcement}). This question is to be attacked using the concept of \textit{mixed strategy} in a game. A mixed strategy is a probability distribution over available actions for an agent, so similarly to the $\varepsilon$-greedy methods in Reinforcement Learning we allow for some actions to be randomly choosen in order to explore further possibilities.

@@

The chosen actions will be compared to classic on-line decision making 
\subsubsection{Internal Validation: Reproducibility}
For the results obtained by following the methodology described in previous sections be reproducible, the causal model used to simulate an environment must be fully specified by describing the probabilistic parameters that control it. The initial distributions given to the decision maker to encode his beliefs must be shown as well with the updating criteria used when information from the environment becomes available. 
\subsubsection{External Validation: Different problems}
For the proposed method to be tested in different decision problems, it is required that the correspondent causal model satisfies all of the conditions previously described, and any derived algorithm will be designed in such a way that the characteristics of the causal model are an external input. In this way, if an external user is able to give a causal model, the method is expected to work independent of the particularities of the causal model which is being tested.
\subsubsection{Concurrent Validation}
Concurrent validation will be tested by comparing the performance obtained by the proposed method against the theoretical optima.
\subsection{Working plan}
Concrete steps towards solving the problem:
\begin{itemize}
\item Literature review, understanding the problem, considering several models and possible solutions.
\item Formalization of a Causal Decision Problem when causal relations are encoded into a Causal Graphical Model.
\item Studying implications of the rationality assumption for decisiom making under uncertainty in causal environments.
\item Propose a representation of beliefs that allow causal structure to be considered. Three subcases considered
\item Updating criteria for beliefs.
\item Discovery of causal structure.
\item Considering interventions in more than one variable.

\end{itemize}
\subsubsection{Publications plan}
\begin{itemize}
\item \textbf{Year 1:} Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML / IJCAI / AAMAS 2018. July 15 2018, Stockholm, Sweden. This can be confirmed at \url{https://sites.google.com/site/faim18wscausalml/program-schedule}. The paper can be found at \url{https://arxiv.org/abs/1807.01268}.
\item \textbf{Year 2:} Conference Paper
		\begin{itemize}
			\item When a causal model is unknown, how to encode uncertainties about causal structure in a probabilistic distribution? Bayesian Nonparametric prior that explicitly represents causal structure.
		\end{itemize}
\item \textbf{Year 2:} Conference Paper
	\begin{itemize}
		\item Using game-theoretical aspects for balancing exporation and exploitation for causal discovery. 
	\end{itemize}
\item \textbf{Year 3:} Journal Paper
		\begin{itemize}
			\item Aim: Theoretical paper of results for on-line causal learning.
		\end{itemize}
\end{itemize}

@@

\newpage
\includepdf[pages=-]{paper_arxiv.pdf}

>>>>

% Title page
\input{gfrontpage}
%\doublespacing
\singlespacing
\tableofcontents
\newpage
\begin{abstract}
We consider decision problems under uncertainty where the options available to a decision maker and the resulting outcome are related through a causal mechanism which is unknown to the decision maker, although he is aware of the causal nature of his environment. We study how a decision maker can learn about this causal mechanism through sequential decision making as well as using current causal knowledge inside each round in order to make better choices had he not considered causal knowledge. We propose a decision making procedure in which an agent holds \textit{beliefs} about her environment which are used to make a choice and then are updated using the observed outcome. As proof of concept, we present an implementation of this causal decision making model and apply it to a simple problem. We show that the model achieves a performance similar to the classic Q-learning while it also acquires a causal model of the environment. 
\end{abstract}
\pagenumbering{arabic}
\section{Introduction}
\indent A fundamental part of intelligent reasoning is being able to make decisions under uncertain conditions (\cite{danks2014unifying}, \cite{lake2017building}, \cite{pearlwhy}). Uncertain conditions as when that a set of choices is made available to a decision maker he can not predict the precise consequence of his decision. For this reason, the decision maker must take into account the stochastic relation between actions\footnote{Actions, options and choices to be used equivalently.} and consequences. From the von Neumann-Morgenstern decision making theory we know that if a rational\footnote{To be explained in full detail later.} decision maker knows this stochastic relation, then he must\textit{If rationality is assumed} choose the action that maximizes the \textit{expected utility} of a \textit{utility function} whose existence is guaranteed (\cite{von1944theory}). On the other hand, if a rational decision maker does not know the probabilities of the ocurrence of certain outcomes, or consequences, given his actions, then he must encode his personal, subjective, quantifications of any uncertainties in a probability distribution $p$ and choose the action that maximizes expected utility with respect to this probability distribution(\cite{savage1954the}, \cite{bernardo2000bayesian}, \cite{gilboa2009decision}). If the current \textit{knowledge} of a decision maker is not enough to make good choices, then the decision maker can \textit{learn} from the environment by interacting with it in order to update his quantifications of uncertainty (\cite{bernardo2000bayesian}, \cite{peterson2017introduction}).

@@

Learning by interaction has been extensively studied by computer scientists using the Reinforcement Learning (RL) setting (\cite{sutton1998reinforcement}), but the most common used techniques  in this field are purely associative and do not consider any high-level structure of the environment beyond what is expressable in a Markov Decision Process (\cite{garnelo2016towards}). A particular case of a \textit{higher level structure}; i.e., beyond associative patterns, is the case of \textit{causal} structure. A causal structure encondes a series of \textit{cause-effect} relations between events and knowing such relations allows a decision maker to add extra knowledge into the uncertainty of his environment and also allows to plan ahead his actions since he can predict what a certain action will cause (\cite{spirtes2000causation}, \cite{pearl2018theoretical}).\\
\\
Causal reasoning is found at the very core of human reasoning, since it allows us to manipulate our environment and being able to predict effects of a given action (\cite{spirtes2000causation}). Causal claims also support reasoning of the form \textit{if...hadn't occurred then...wouldn't have happened}. It is argued by \cite{pearl2018why} that imagining alternative scenarios is fundamental for humans to make causal reasoning.  Since human beings are known to learn causal models in sequential decision making processes (\cite{sloman2006causal}, \cite{nichols2007decision}, \cite{meder2010observing}, \cite{hagmayer2013repeated}, \cite{danks2014unifying}), and even though this learning is not perfect (\cite{rottman2014reasoning}), our hypothesis is that an autonomous agent can learn and use causal information while interacting with an uncertain environment which is governed by a fixed \textit{causal mechanism} which is unknown to the agent.

@@

Learning a causal model of a certain environment would allow transferability of knowledge into similar domains because the underlying cause-effect structure has been captured and could later be used.
\subsection{Outline}
In the Section \ref{Main_Concepts} the main concepts that serve as a foundation for this work will be described; in particular, we will describe in detail what causality means. Since this research proposal asumes that a causal model governs an environment, a precise definition of causality will be given along with the mathematical machinery to perform causal inference. Also, since the entity that will be facing causal environments is a \textit{rational} decision maker we will precisely define, axiomatically, what does rationality means and which are the main theories of rational decision making. At the end of the section, we will describe a mathematical theory which is suitable to modelling interaction between rational decision makers since interactive learning is considered.

@@

	Notice that since the graph is finite, there exist some nodes that does not have causes, which are called \textit{exogenous}. If an $A$ event is caused by some other event, then we say it is \textit{endogenous} and we denote the set of its causes as $Pa(A)$. It is proven in \cite{kiiveri1984recursive} that at least one exogenous node exists in a causal graph.

@@

		\subsubsection{Extensions}
		Given the Limitations addressed in the previous section, some possible extensions could be:
		\begin{itemize}
		\item Consider longer games.
		\item Consider more players than an agent and Nature.
		\item Consider objectives/intentions for Nature.
		\item Studying game-theoretical properties of the model such as equilibria in terms of the causal structure.
		\end{itemize}
\newpage
\section{Related Work}{\label{Related_Work}}
Reinforcement Learning (RL) has been the standard setting for decision-making and learning by interaction. In this setting it is assumed that a Markov Decision Process (MDP) is able to capture the effects of the agent's actions in the environment. An MDP is composed by a set of states, a transition function and a reward function, where an agent's action moves the state of the environment according to the transition function and obtaining a reward in such process.

@@

\newpage
\section{Methodology}
In this section we describe the steps that are to be followed in order to answer the Hypothesis.
\subsection{A guiding principle for a Causal Decision Problem}
We propose the following principle as a normative guide for how causal information must be used in order to find a solution to a decision problem where the preferences of the decision maker are assumed to be rational. We consider only the case where the decision maker's utility is given by the outcome of a $\{0,1 \}$-binary variable.
\begin{prop}
Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Let $Y$ a variable in $\mathcal{G}$ the variable whose realizations correspond to the consequences of the decision problem once the agent has chosen an action $a \in \mathcal{A}$ and assume that $Y$ only takes values in the set $\{ 0,1\}$ where $1$ is a more desired outcome for the agent. Then, the $\textit{solution}$ for this decision problem is given by the action $a^\ast \in \mathcal{A}$ that satisfies:
\[ P(Y=1 | do(a^\ast) \geq P(Y=1 | do(a)) \textrm{ for all } a \in \mathcal{A}. \]
\end{prop}
\subsection{Modelling a Decision Problem under Uncertainty as a Game}
Consider a decision problem under uncertainty $(\mathcal{A},\mathcal{E},\mathcal{C},\succeq)$ where the available actions $\mathcal{A}$ to a decision maker are causally related to the outcomes $\mathcal{C}$ and this causal relations are captured by a Causal Graphical Model $\mathcal{G}$. The Causal model is assumed to remain fixed and unknown to the decision maker, who is assumed to be rational and aware of the causal nature of the decision problem he faces.

@@

If the maximum number of variables is assumed to be known by the decision maker, then the situation simplifies since a decision maker could generate distributions that represent causal relations among variables. For example, a Dirichlet Process (\cite{ferguson1973bayesian}, \cite{ghosal2017fundamentals}) could be used to generate Dirichlet distributions that are used generate Conditional Probability Tables in order to specify a causal model as in the previous case.

@@

\subsection{Balancing exploitationd and exploration}
When the model is to be discovered, such as in the general case mentioned before, an agent must choose also between keep making an intervention in a variable that is known to cause a desired output or to explore for new variables that could bring even better rewards, but at the cost of not having best performance while looking out for them; this is known as the exploration vs exploitation dilemma (\cite{sutton1998reinforcement}). This question is to be attacked using the concept of \textit{mixed strategy} in a game. A mixed strategy is a probability distribution over available actions for an agent, in this way, similar to the $\varepsilon$-greedy methods in Reinforcement Learning, we allow for some actions to be randomly choosen in order to explore further possibilities.

@@

The chosen actions will be compared to classic on-line decision making 
\subsubsection{Internal Validation: Reproducibility}
For the results obtained by following the methodology described in previous sections be reproducible, the causal model used to simulate an environment must be fully specified by describing the probabilistic parameters that control it. The initial distributions given to the decision maker to encode his beliefs must be shown as well with the updating criteria used when information from the environment becomes available. 
\subsubsection{External Validation: Different problems}
For the proposed method to be tested in different decision problems, it is required that the correspondent causal model satisfies all of the conditions previously described, and any derived algorithm will be designed in such a way that the characteristics of the causal model are an external input. In this way, if an external user is able to give a causal model, the method is expected to work independent of the particularities of the causal model which is being tested.
\subsubsection{Concurrent Validation}
Concurrent validation will be tested by comparing the performance obtained by the proposed method against the theoretical optima.
\subsection{Working plan}
Concrete steps towards solving the problem:
\begin{itemize}
\item Literature review, understanding the problem, considering several models and possible solutions.
\item Studying implications of the rationality assumption for decision making under uncertainty in causal environments.
\item Formalization of a Causal Decision Problem: a decision problem under uncertainty where causal relations are encoded into a Causal Graphical Model. Propose a solution using the rationality assumption.
\item Propose a representation of beliefs that allow causal structure to be considered. Three subcases considered
\item Propose and evaluate an updating criteria for causal beliefs that can be updated according to new information.
\item Study the problem of on-line causal discovery.
\item Considering interventions in more than one variable.

\end{itemize}
\subsubsection{Publications plan}
\begin{itemize}
\item \textbf{Year 1:} Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for poster presentation at the \textbf{CausalML 2018 Workshop at ICML 2018}. July 15 2018, Stockholm, Sweden. This can be confirmed at \url{https://sites.google.com/site/faim18wscausalml/program-schedule}. The paper can be found at \url{https://arxiv.org/abs/1807.01268}.
\item \textbf{Year 2:} Conference Paper
		\begin{itemize}
		\item The implications of Causal Relations for Decision Making: a general principle for Decision under Uncertainty.
		\end{itemize}
\item \textbf{Year 2:} Conference Paper
	\begin{itemize}
	\item When a causal model is unknown, how to encode uncertainties about causal structure in a probabilistic distribution? Bayesian Nonparametric prior that explicitly represents causal structure.
	\end{itemize}
\item \textbf{Year 3:} Journal Paper
		\begin{itemize}
			\item Aim: Theoretical paper of results for on-line causal learning.
		\end{itemize}
\end{itemize}

@@
